---
title: "Scaled logit model with inclusion of unexposed population"
author: "Arseniy Khvorov"
date: "10/12/2019"
documentclass: article
geometry: 
  margin=1in
output: 
  bookdown::pdf_document2:
    keep_tex: yes
    toc: no
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

All relevant code is at github.com/khvorov45/sclr-lowbase

# Context

The aim is to investigate the ralationship between a covariate (e.g. antibody titre) and a binary outcome (e.g. infection status).

If the sample contains people unexposed to the pathogen (and therefore at no risk of the outcome), including them into the analysis will not bias the estimates under the scaled logit model

$$
\begin{gathered}
P(Y=1|E=1) = \frac{\lambda}{1+\text{exp}(\beta_0+\beta_T T)} \\
P(Y=1|E=0) = 0 \\
P(Y=1) = \frac{P(E=1)\lambda}{1+\text{exp}(\beta_0+\beta_T T)} = \frac{\lambda^*}{1+\text{exp}(\beta_0+\beta_T T)}
\end{gathered}
$$

Where $\lambda^* = \lambda P(E=1)$.

Including the unexposed into the analysis has the expected effect of lowering the baseline estimate by the probabiity of exposure. The infection curve can be expected to have a lower top plateau but the same logistic slope and intercept. The protection curve can be expected to be unaffected since it only depends on the $\beta$ parameters.

However, including the unexposed into the analysis has an effect of the expected standard errors of the parameter estimates. 

# Effect of the unexposed on SE of parameter estimates

To investigate the effect of including the unexposed population into the analysis on the standard errors of the parameter estimates, I simulated data from the model

$$
\begin{gathered}
P(Y=1) = \frac{P(E=1)\text{exp}(\theta)}{(1+\text{exp}(\theta))(1+\text{exp}(\beta_0+\beta_T T))} \\
\theta = \text{log}(\frac{\lambda}{1-\lambda}) \\
\lambda = 0.5 \quad \beta_0 = -5 \quad \beta_T = 1.5
\end{gathered}
$$

at different values of $P(E=1)$.

I then fit the scaled logit model using maximum likelihood to all data (general population, unexposed included) and the subset with just the exposed (unexposed excluded). The results of 10,000 simulations at each value of $P(E=1)$ from 0.1 to 1 are in Figure \@ref(fig:plot-10000sims).

(ref:plot-10000sims) The results of 10,000 simulations at each parameter combination. Points represent values at which simulations were performed. Lines are colored based on the estimated term they belong to. The solid lines show estimated mean (left panel) and mean standard error (right panel) of estimates obtained from fitting models to the exposed population (i.e. unexposed excluded). The dashed lines show the same for the general population (i.e. unexposed included).

```{r plot-10000sims, echo=FALSE, fig.align='center', fig.cap='(ref:plot-10000sims)'}
knitr::include_graphics("sim-plot/plot-10000sims.pdf")
```

As expected, including the unexposed into the analysos has the effect of lowering the estimated baseline probability but has no appritiable effect on the expected estimates of the other parameters.

Including the unexposed into the anlalysis also has the effect of increasing the expected standard errors of the $\beta$ parameters (especially the intercept $\beta_0$) by 5-10% when the exposed proportion is less then 50%.

# Conclusion

Including the unexposed into the analysis offers no benefits and has the detriment of increasing estimate error (despite the fact that there is more data with the unexposed included). If there is a good way to isolate the unexposed subjects in the analysis, those observations should be excluded.
